---
title: "EDS212_Resource_Document"
format: html
editor: visual
---

# Quarto

```{r}
#| fig-height: 5 
#| fig-width : 4
#these set the width of figures in output
library(here)
library(tidyverse)
library(janitor)
library(dplyr)
```

# Terminal

The terminal in R studio is the same as the terminal outside r studio.

Navigate to a folder that is version controlled and you can use git commands.

-   Git Add -A

-   Git commit -m "important message"

-   git push

-   git status

pwd

-   tells you where you are

cd ..

-   will take you back up one

ls

-   list all files

cd

-   sets working directory

# Pipe Operator

%\>% pipe operator or \|\>

-   basically means and then, allows you to write code in the order you think about it.

-   command shift m- shortcut

# Working With Files

## Here

Here Package: much better option than using working directory

## Importing and Exporting Files + Data Tidying

add file paths starting from here location

```{r}
# import and fix the names
blood_lead <- read_csv(here("data","stl_blood_lead.csv" )) %>% clean_names() #automatically makes names lowercase
#make a new column
blood_lead_pov <- mutate(blood_lead, 
                         new_col = poverty_tot/total_pop *100)
#Graph a plot
lead_poverty_plot <- ggplot(data = blood_lead_pov, aes(x = pct_elevated, y = new_col)) + geom_point(color = "pink") +labs(x = "percent of population in poverty", y = "Percent of population with elevated blood levels")
lead_poverty_plot
#save the file in the project
ggsave(plot = lead_poverty_plot, here("figs","lead_plot.png"), width = 4, height = 4)
```
